---
title: "Black Cherry Tree Logging App"
author: "Lucas Qualmann"
date: "10/15/2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Logging Black Cherry Trees

- Are you bad at estimating how much wood you'll get from chopping down black cherry trees?
- Do you always park your truck to close to the tree you're cutting down?
- Would you just like to do your job without having to think so hard?

Our app is your solution!

## How it Works

Easily estimate the voume and height of your black cherry tree with one simple measurement!

- Measure the diameter of the tree in inches 4 ft 6 in off the ground.
- Plug that measurement into the slider on the app.
- See an estimate for the tree's volume and height.

## Sample Volume Plot
```{r echo=FALSE}
modelV <- lm(Volume ~ Girth, data = trees)
predV <- round(predict(modelV, newdata = data.frame(Girth = mean(trees$Girth))),
               2)
plot(trees$Girth, trees$Volume, pch = 19, col = "blue", cex = 1.5, 
     xlab = "Diameter (inches)", ylab = "Volume (cubic feet)", 
     main = "Black Cherry Tree Volume vs Diameter", xlim = c(8, 21), 
     ylim = c(0, 80))
abline(modelV, col = "orange", lwd = 3)
abline(v = mean(trees$Girth), lwd = 2, lty = 2, col = "red")
abline(h = predV, lwd = 2, lty = 2, col = "red")
```

Predicted Volume (cubic feet):  `r predV`

## Data Used

The trees dataset in R contains 31 observations of felled black cherry trees with 3 variables: girth (diameter in inches), height (in feet), and volume (in cubic feet).  See a summary of the data below.

```{r cars, echo = TRUE, comment=""}
summary(trees)
```



